Using News' Title and Body Text Features to Detect Fake News: a Viability Study

### 2019-09-04 FakeNewsAnalysis/paperLatex/paperFelipeNevesDataScience.tex
improved plot and text

We also tried increasing the number of features in order to try to improve the performance,
specially in the expanded dataset that had a very poor performance with the original features.
In these tests we also switched the classifier for the Random Forest classifier due to the
SVM being to slow to calculate with too many features and entries. Previous tests had already
suggested that the classifier had little impact in the overall performance.
The results are shown in the Table \ref{table:od_ef} and Table \ref{table:od_of}.

@@

TODO: plot new graph

@@

@@

\section*{Acknowledgement}
I would like to thank Prof. Luiz Celso Gomes Junior from UTFPR for the support
provided during the development of this project.

>>>>

We also tried increasing the number of features in order to try to improve the performance,
specially in the expanded dataset that had a very poor performance with the original features.
In these tests we also switched the classifier for the Random Forest classifier due to the
SVM being to slow to calculate with too many features and entries.

Previous tests had already suggested that the classifier had little impact in the overall performance.
The results are shown in the Table \ref{table:od_ef} and Table \ref{table:od_of}.

@@

For the purpose of being able to better distinguish the results on each step we show
in Figure \ref{fig:results} a comparison between all the results side by side.

\begin{figure}[htbp]
\centerline{
    \includegraphics[width=0.5\textwidth]
    {content/results.png}
    }
\caption{Results comparison}
\label{fig:results}
\end{figure}

@@

As it is clear to see in Figure \ref{fig:results} it is clear that the performance over
this bigger dataset was significantly worse than in the original paper \cite{horne_2017},
however it is important to understand why whe had such difference.

One very obvious reason for the divergence of the results of the classifier for the original
dataset is that the algorithms, libraries and classifiers utilized in our implementation.
This means that even though the author shared the dataset \cite{horne_data_2017}, as they
did not share their implementation and it makes it very hard to reproduce the same results.

In the field of Data Science reproducibility is a very important issue because due to the
problem that finding reliable and unbiased datasets represent we often find the need to
test and validate solutions proposed in other peoples' works. And in cases where either
the source code or the datasets are not available it becomes near impossible to do
\cite{peng_2011}\cite{peng_2015}.

In this case the main problem with the original dataset  is that no only it had only 150
entires, but also that it had only one subject, the 2016 American Elections, and what this
means in terms of the features and classifiers is that the reliable news and fake news will
share many similar characteristics among the items in their groups, an thus making it easier
to achieve a better accuracy on the classifier output. It is easier to separate groups when
their items are very similar inside each group, and share the same differences between items
of the two groups.

As we can see on Figure \ref{fig:titleNN}, some features did not have a  different distribution
over the fake and reliable news, meaning that a classifier would not be able to separate these
classes based on this feature only. Using more features helped overcome this problem but some
features may actually decrease the overall performance.

\begin{figure}[htbp]
\centerline{
    \includegraphics[width=0.5\textwidth]
    {content/titleFK.png}
    }
\caption{FK index distribution over titles}
\label{fig:titleNN}
\end{figure}

This also brings the problem that the features were chosen among the ones with the highest
statistical significance in the original dataset, and this significance may not hold up
in a different dataset such as ours. Another issue is that as the original dataset had the
news manually chosen it is more susceptible to bias, and the fake news chosen were more
explicit, with more distinguishable features compared to reliable news, than the ones in
our expanded dataset.

@@

The main point that we can drawn from this analysis is that the same feature set may not be
efficient for classifying two different datasets. In this case the characteristics of the two
datasets should be compared to see if they are similar enough. Therefore this approach of
trying to use few but very specific features does not work well in the general case, because
any change in the dataset may reduce the classification performance. To actually improve overall
performance the best approach is to choose a different and wider set of features, and avoid choosing
them based on a specific dataset, this way we can avoid the overfitting of the classification,
where new inputs will have a worse performance due to being just slightly difference from the
initial dataset.

However only using more features is good but may have some drawbacks in regarding the processing
speed that will be increased. Also, if the features set contains features that are not relevant
to the analysis the performance may actually decrease due to containing inputs that cannot be
separated by the classifier, such as what happened with the performance of out classification using
the original and expanded features over the titles of the original dataset, where the performance
dropped from 72.67\% with the original features to 71.33\% when using the expanded features, as shown
on Table \ref{table:od_of} and Table \ref{table:od_ef} respectively.

Another factor that can influence the performance is the choice of the classifier itself.
The SVM is not appropriate for non-normalized data, because each input will influence the others
and therefore the results may be good even though some features are having little impact in the
classification. A more appropriate solution in this cases would be either normalize all the data
or to utilize another type of classifier such as Random Forest, where the inputs do not influence
each other and therefore requires no normalization. Also, when working with larger datasets the
computational power required has a great impact, and the Random Forest algorithm resulted in
significantly lower processing times, to the order of 100 times. But this of course cannot be
generalized as the number of entries and configuration of each classifier can heavily impact
on their speed.

\section*{Acknowledgement}

I would like to thank Prof. Luiz Celso Gomes Junior from UTFPR for the support
provided during the development of this project. It is also important to commend
the work of all the people that collaborate developing free Data Science libraries
and other resources that helps making Data Science more accessible, and without
whom this work would not be possible.
